---
title: "Input, Output, and Spectrum Conversion"
author: "Xiaotao Shen"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Input, Output, and Spectrum Conversion}
  %\VignettePackage{masstools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Read example files

```{r}
library(masstools)
```

The package includes small example MGF and mzXML files.

```{r}
mgf_file <- system.file("extdata", "example.mgf", package = "masstools")
mzxml_file <- system.file("extdata", "example.mzXML", package = "masstools")

mgf_file
mzxml_file
```

### Read MGF

```{r}
mgf_data <- suppressWarnings(read_mgf(mgf_file))
length(mgf_data)
str(mgf_data[[1]], max.level = 1)
```

### Read mzXML

```{r}
mzxml_data <- read_mzxml(mzxml_file)
length(mzxml_data)
str(mzxml_data[[1]], max.level = 1)
```

## Convert between data frames and `Spectra`

Many workflows start from a simple two-column spectrum table.

```{r}
spectrum_df <- data.frame(
  mz = c(100, 120, 150),
  intensity = c(10, 40, 100)
)

spectrum_df
```

Convert it to a `Spectra` object:

```{r}
spec <- df_to_spectra(
  spectrum_df,
  precursor_mz = 150.0,
  precursor_rt = 25.0,
  spectrum_id = "example-spectrum"
)

spec
```

And convert back to a plain data frame:

```{r}
spectra_to_df(spec)
```

## Match feature tables

`match_mz_rt()` is useful when comparing peak tables from two runs or two
processing steps.

```{r}
table_1 <- data.frame(
  mz = c(100, 150, 200),
  rt = c(10, 20, 30)
)

table_2 <- data.frame(
  mz = c(100.0002, 149.9998, 210),
  rt = c(10.3, 19.8, 35)
)

match_result <- match_mz_rt(
  table_1,
  table_2,
  mz.tol = 10,
  rt.tol = 1,
  rt.error.type = "abs"
)

match_result
```

If one feature matches multiple candidates, use `keep_best_match()` to keep the best
match.

```{r}
duplicated_result <- data.frame(
  Index1 = c(1, 1),
  Index2 = c(1, 2),
  mz1 = c(100, 100),
  mz2 = c(100, 100.00005),
  `mz error` = c(0, 0.5),
  rt1 = c(10, 10),
  rt2 = c(10.4, 10.1),
  `rt error` = c(0.4, 0.1),
  check.names = FALSE
)

keep_best_match(duplicated_result, according.to = "mz.error")
keep_best_match(duplicated_result, according.to = "rt.error")
```

## Package documentation

After installation, users can open the vignette index with:

```{r eval = FALSE}
browseVignettes("masstools")
```
