---
title: "Quick Start"
author: "Xiaotao Shen"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignettePackage{masstools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Overview

`masstools` provides small, reusable utilities for mass spectrometry workflows.
It is especially useful for:

- chemical formula parsing and validation
- MS/MS spectrum matching and scoring
- reading common MS data formats
- converting between simple spectrum tables and Bioconductor `Spectra` objects

```{r}
library(masstools)
```

## Formula utilities

Start with a chemical formula and inspect, validate, and calculate masses.

```{r}
check_chemical_formula("C6H12O6")
parse_chemical_formula("C6H12O6")
calculate_mass("C6H12O6", which = "exact_mass")
combine_formula_adduct("C9H11NO2", "M+H")
```

`convert_precursor_mz2accurate_mass()` is useful when the observed precursor
contains a supported adduct.

```{r}
convert_precursor_mz2accurate_mass(180.0634, "(M+H)+")
```

## Spectrum matching

Create two small MS/MS spectra and compare them.

```{r}
exp_spectrum <- data.frame(
  mz = c(100, 120, 150, 180),
  intensity = c(10, 40, 100, 20)
)

lib_spectrum <- data.frame(
  mz = c(100, 120.0005, 149.999, 190),
  intensity = c(15, 35, 95, 10)
)

ms2_match(exp_spectrum, lib_spectrum, ppm.tol = 10)
calculate_dot_product(c(1, 2, 3), c(1, 2, 3))
calculate_spectra_match_score(exp_spectrum, lib_spectrum, ppm.tol = 10)
```

You can also visualize one or two spectra.

```{r}
plot_ms2(exp_spectrum)
plot_ms2(exp_spectrum, lib_spectrum)
```

## Feature matching by m/z and retention time

For feature tables, `match_mz_rt()` matches entries using both `mz` and `rt`.

```{r}
features_1 <- data.frame(mz = c(100, 200, 300), rt = c(30, 60, 90))
features_2 <- data.frame(mz = c(100.0002, 199.9995, 305), rt = c(31, 58, 95))

result <- match_mz_rt(features_1, features_2, mz.tol = 10, rt.tol = 5, rt.error.type = "abs")
result
keep_best_match(result, according.to = "mz.error")
```

## Read bundled example files

The package ships with small example files in `extdata`.

```{r}
mgf_file <- system.file("extdata", "example.mgf", package = "masstools")
mzxml_file <- system.file("extdata", "example.mzXML", package = "masstools")

mgf_data <- suppressWarnings(read_mgf(mgf_file))
mzxml_data <- read_mzxml(mzxml_file)

length(mgf_data)
length(mzxml_data)
names(mgf_data[[1]])
```

## Next steps

After installation, users can browse the available articles with:

```{r eval = FALSE}
browseVignettes("masstools")
```
