---
title: "Formula and Spectra Workflows"
author: "Xiaotao Shen"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Formula and Spectra Workflows}
  %\VignettePackage{masstools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Chemical formulas

`masstools` includes several helpers for validating and transforming chemical
formulas.

```{r}
library(masstools)
```

### Validate and parse

```{r}
check_chemical_element("NaCl")
check_chemical_formula("H2O")
check_chemical_formula("2H")
parse_chemical_formula("C10H14N2")
```

### Exact and average mass

```{r}
calculate_mass("H2O", which = "exact_mass")
calculate_mass("H2O", which = "average_mass")
```

### Add adducts or neutral losses

```{r}
combine_formula_adduct("C9H11NO2", "M+H")
combine_formula_adduct("C9H11NO2", "M+CH3COOH")
combine_formula_adduct("C9H11NO2", "M-H2O+H")
```

### From precursor m/z to neutral mass

```{r}
convert_precursor_mz2accurate_mass(181.0707, "(M+H)+")
convert_precursor_mz2accurate_mass(179.0561, "(M-H)-")
```

## MS/MS similarity

### Remove redundant fragment peaks

```{r}
spec <- data.frame(
  mz = c(100, 100.0004, 120, 150),
  intensity = c(10, 15, 50, 100)
)

remove_noise(spec, ppm.ms2match = 10)
```

### Match two spectra

```{r}
exp_spec <- data.frame(
  mz = c(100, 120, 150, 180),
  intensity = c(10, 20, 80, 30)
)

lib_spec <- data.frame(
  mz = c(100, 120.0003, 151, 180),
  intensity = c(12, 18, 20, 25)
)

match_tbl <- ms2_match(exp_spec, lib_spec, ppm.tol = 10)
match_tbl
```

### Score similarity

```{r}
calculate_dot_product(match_tbl$Exp.intensity, match_tbl$Lib.intensity)
calculate_spectra_match_score(exp_spec, lib_spec, ppm.tol = 10)
```

### Plot spectra

```{r}
plot_ms2(exp_spec)
plot_ms2(exp_spec, lib_spec, ppm.tol = 10)
```

## Deprecated compatibility functions

Backward-compatible wrappers are still available, but new code should use:

- `calculate_dot_product()`
- `ms2_match()`
- `remove_noise()`
